---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# Overview

The wedge package provides functionality for working with the exterior
calculus.  It includes cross products and wedge products and a variety
of use-cases.  A vignette is provided in the package.

# Installation

You can install the released version of wedge from [CRAN](https://CRAN.R-project.org) with:

```{r}
# install.packages("wedge")  # uncomment this to use the package
library("wedge")
```

# The ```wedge``` package in use

The package has two main classes of objects, ```kform``` and
```ktensor```.  We may define a $k$-tensor as follows

```{r ktensor}
KT <- as.ktensor(cbind(1:4,3:5),1:4)
KT
```

We can coerce ```KT``` to a function and then evaluate it:

```{r evaluatektensor}
KT <- as.ktensor(cbind(1:4,2:5),1:4)
f <- as.function(KT)
E <- matrix(rnorm(10),5,2)
f(E)
```

Cross products are implemented:

```{r crossprod}
KT %X% KT
```


## Alternating forms

An alternating form (or $k$-form) is an antisymmetric $k$-tensor; the
package can convert a general $k$-tensor to alternating form using the
```Alt()``` function:


```{r showalt}
Alt(KT)
```

However, the package provides a bespoke and efficient representation
for $k$-forms as objects with class ```kform```.  Such objects may be
created using the ```as.kform()``` function:

```{r example}

M <- matrix(c(4,2,3,1,2,4),2,3,byrow=TRUE)
M
KF <- as.kform(M,c(1,5))
KF
```

We may coerce ```KF``` to functional form:

```{r e2}
f <- as.function(KF)
E <- matrix(rnorm(12),4,3)
f(E)
```

# The wedge product

The wedge product of two $k$-forms is implemented as ```%^%``` or
```wedge()```:

```{r definekf2}
KF2 <- kform_general(6:9,2,1:6)
KF2
KF %^% KF2
```

The package can accommodate a number of results from the exterior
calculus such as elementary forms:

```{r dxdyxz}
dx <- as.kform(1)
dy <- as.kform(2)
dz <- as.kform(3)
dx %^% dy %^% dz  # element of volume 
```

A number of useful functions from the exterior calculus are provided,
such as the gradient of a scalar function:

```{r grad}
grad(1:6)
```

The package takes the leg-work out of the exterior calculus:

```{r legwork}
grad(1:4) %^% grad(1:6)
```