---
title: "The `inner()` function in the `stokes` package"
author: "Robin K. S. Hankin"
output: html_document
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{inner}
  %\usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("stokes")
library("emulator")
set.seed(0)
```

# The `inner()` function in the `stokes` package


```{r}
inner
```

Function `inner()` returns the inner product corresponding to a
matrix.

The inner product of two vectors $\mathbf{x}$ and $\mathbf{y}$ is
usually written $\left\langle\mathbf{x},\mathbf{y}\right\rangle$ or
$\mathbf{x}\cdot\mathbf{y}$, but the most general form would be
$\mathbf{x}^TM\mathbf{y}$ where $M$ is a matrix.  Noting that inner
products are multilinear, that is $\left\langle\mathbf{x},a\mathbf{y}+b\mathbf{z}\right\rangle=a\left\langle\mathbf{x},\mathbf{y}\right\rangle
+ b\left\langle\mathbf{x},\mathbf{z}\right\rangle$ and $\left\langle
a\mathbf{x} +
b\mathbf{y},\mathbf{z}\right\rangle=a\left\langle\mathbf{x},\mathbf{z}\right\rangle
+ b\left\langle\mathbf{y},\mathbf{z}\right\rangle$ we see that the
inner product is indeed a multilinear map, that is, a tensor.

We can start with the simplest inner product, the identity matrix:

```{r}
inner(diag(7))
```

Note how the rows of the tensor appear in arbitrary order.  Verify:

```{r}
x <- rnorm(7)
y <- rnorm(7)
V <- cbind(x,y)
c(as.function(inner(diag(7)))(V),sum(x*y))  # should match
```

A more stringent test would be to use a general matrix:

```{r}
M <- matrix(rnorm(49),7,7)
f <- as.function(inner(M))
c(f(V),quad.3form(M,x,y)) # should match
```

(function `emulator::quad.3form()` evaluates matrix products
efficiently; `quad.form(M,x,y)` returns $x^TMy$).  Of course, we would
expect `inner()` to be a homomorphism:


```{r}
M1 <- matrix(rnorm(49),7,7)
M2 <- matrix(rnorm(49),7,7)
g <- as.function(inner(M1+M2))
c(g(V),quad.3form(M1+M2,x,y)) # should match
```

Now, if the matrix is symmetric the corresponding inner product should also be symmetric:

```{r}
h <- as.function(inner(M1 + t(M1)))
c(h(V), h(V[,2:1]))  # should match
```

Also positive-definite matrix should return a positive quadratic form:

```{r}
M3 <- crossprod(matrix(rnorm(56),8,7))  # 7x7 pos-def matrix
as.function(inner(M3))(kronecker(rnorm(7),t(c(1,1))))>0  # should be TRUE
```

## Alternating forms

The inner product on an antisymmetric matrix should be alternating:

```{r}
jj <- matrix(rpois(49,lambda=3.2),7,7)
M <- jj-t(jj) # M is antisymmetric
f <- as.function(inner(M))
c(f(V),f(V[,2:1])) # differ in sign
```
