---
title: "Exterior calculus with R"
author: "Robin K. S. Hankin"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{The exterior calculus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# The exterior calculus

Ordinary differential calculus may be generalized to
arbitrary-dimensional oriented manifolds using the exterior calculus.
Here I show how the wedge package furnishes functionality for working
with the exterior calculus, and provide numerical verification of a
number of theorems.

Notation follows that of Spivak, and Hubbard and Hubbard.

```{r,echo=FALSE,print=FALSE,results='hide'}
require("wedge",quietly=TRUE,warn.conflicts=FALSE)
require("Deriv",quietly=TRUE,warn.conflicts=FALSE)
```

Recall that a $k$-tensor is a multilinear map $S\colon
V^k\longrightarrow\mathbb{R}$, where $V=\mathbb{R}^n$ is considered as
a vector space; Spivak denotes the space of multilinear maps as
$\mathcal{J}^k(V)$.  Formally, multilinearity means

\[
S\left(v_1,\ldots,av_i,\ldots,v_k\right)=a\cdot S\left(v_1,\ldots,v_i,\ldots,v_k\right)
\]

and

\[S\left(v_1,\ldots,v_i+{v_i}',\ldots,v_k\right)=S\left(v_1,\ldots,v_i,\ldots,x_v\right)+
S\left(v_1,\ldots,{v_i}',\ldots,v_k\right).
\]

where $v_i\in V$.  If $S\in\mathcal{J}^k(V)$ and
$T\in\mathcal{J}^l(V)$, then we may define $S\otimes
T\in\mathcal{J}^{k+l}(V)$ as

\[
S\otimes T\left(v_1,\ldots,v_k,v_{k+1},\ldots,v_{k+l}\right)=
S\left(v_1,\ldots,v_k\right)\cdot
T\left(v_1,\ldots,v_l\right)
\]


Spivak observes that $\mathcal{J}^k(V)$ is spanned by the $n^k$
products of the form

\[
\phi_{i_1}\otimes\phi_{i_2}\otimes\ldots\otimes\phi_{i_k}\qquad
1\leq i_i,i_2,\ldots,i_k\leq n
\]

where $v_1,\ldots,v_k$ is a basis for $V$ and
$\phi_i\left(v_j\right)=\delta_{ij}$; we can therefore write

\[
S=\sum_{1\leq i_1,\ldots,i_k\leq n} a_{i_1\ldots i_k} 
\phi_{i_1}\otimes\ldots\otimes\phi_{i_k}.
\]

The space spanned by such products has a natural representation in R
as an array of dimensions $n\times\cdots\times n$.  If ```A``` is such
an array, then the element ```A[i_1,i_2,...,i_k]``` is the coefficient
of $\phi_{i_1}\otimes\ldots\otimes\phi_{i_k}$.  However, it is more
efficient and conceptually cleaner to consider a *sparse* array.
Considering the case $n=5,k=4$

We can test these identities in R using the idiom furnished by the
wedge package:

```{r}
k <- 4
n <- 5
M <- matrix(c(5,1,1,1, 1,1,2,3, 1,3,4,2),3,4,byrow=TRUE)
M
S <- as.ktensor(M,coeffs= 0.5 + 1:3)
S
```

So $S$ is a 4-tensor, mapping $V^4$ to $\mathbb{R}$, where
$V=\mathbb{R}^5$.  Here we have
$S=1.5\phi_5\otimes\phi_1\otimes\phi_1\otimes\phi_1+2.5\phi_1\otimes\phi_1\otimes\phi_2\otimes\phi_3+3.5\phi_1\otimes\phi_3\otimes\phi_4\otimes\phi_2$.
Note that on some implementations the row order of object ```S``` will
differ from that of ```M```.  We will define $E$ to be a random point
in $V^k$ in terms of a matrix:

```{r}
E <- matrix(rnorm(n*k),n,k)   # A random point in V^k
```

Then

```{r}
f <- as.function(S)
f(E)
```

Testing  multilinearity is straightforward in the package:
```{r}

E1 <- E
E2 <- E
E3 <- E

x1 <- rnorm(n)
x2 <- rnorm(n)
r1 <- rnorm(1)
r2 <- rnorm(1)

E1[,2] <- x1
E2[,2] <- x2
E3[,2] <- r1*x1 + r2*x2
```

Then we can verify the multilinearity of $S$:

```{r}

f <- as.function(S)
c(r1*f(E1) + r2*f(E2), f(E3)) # should match
```


Note that this is *not* equivalent to linearity over $V^{nk}$:


```{r}
E1 <- matrix(rnorm(n*k),n,k)
E2 <- matrix(rnorm(n*k),n,k)
c(f(E1+E2),f(E1)+f(E2))
```

Given two k-tensor objects $S,T$ we can form the cross product
$S\otimes T$, defined as

\[
S\otimes T\left(v_1,\ldots,v_k,v_{k+1},\ldots, v_{k+l}\right)=
  S\left(v_1,\ldots v_k\right)\cdot T\left(v_{k+1},\ldots
  v_{k+l}\right)
\]


```{r}
S1 <- ktensor(spray(cbind(1:3,2:4),1:3))
S2 <- as.ktensor(matrix(1:6,2,3))
S1
S2
```

The R idiom for $S1\otimes S2$ would be ```cross()```, or ```%X%```:


```{r}
cross(S1,S2)
```

Then, for example:

```{r}
as.function(cross(S1,S2))(matrix(rnorm(30),6,5))
```

# Alternating forms

An alternating form is a multilinear map $T$ satisfying

\[
\mathrm{T}\left(v_1,\ldots,v_i,\ldots,v_j,\ldots,v_k\right)=
    -\mathrm{T}\left(v_1,\ldots,v_j,\ldots,v_i,\ldots,v_k\right)
\]

(or, equivalently,
$\mathrm{T}\left(v_1,\ldots,v_i,\ldots,v_i,\ldots,v_k\right)= 0$).  We
write $\Lambda^k(V)$ for the space of all alternating multilinear maps
from $V^k$ to $\mathbb{R}$.  Spivak gives
$\operatorname{Alt}\colon\mathcal{J}^k(V)\longrightarrow\Lambda^k(V)$
defined by

\[\operatorname{Alt}(T)\left(v_1,\ldots,v_k\right)=
    \frac{1}{k!}\sum_{\sigma\in S_k}\operatorname{sgn}(\sigma)\cdot
    T\left(v_{\sigma(1)},\ldots,v_{\sigma(k)}\right)
\]

where the sum ranges over all permutations of
$\left[n\right]=\left\{1,2,\ldots,n\right\}$ and
$\operatorname{sgn}(\sigma)\in\pm 1$ is the sign of the permutation.
If $T\in\mathcal{J}^k(V)$ and $\omega\in\Lambda^k(V)$, it is
straightforward to prove that $\operatorname{Alt}(T)\in\Lambda^k(V)$,
$\operatorname{Alt}\left(\operatorname{Alt}\left(T\right)\right)=\operatorname{Alt}\left(T\right)$,
and $\operatorname{Alt}\left(\omega\right)=\omega$.

In the wedge package, this is effected by the ```Alt()``` function:


```{r}
S1
Alt(S1)
```

And verifying that it is in fact alternating is straightforward:

```{r}
E <- matrix(rnorm(8),4,2)
Erev <- E[,2:1]
as.function(Alt(S1))(E) + as.function(Alt(S1))(Erev)  # should be zero
```

However, we can see that this form for alternating tensors (here
called $k$-forms) is inefficient and highly redundant.  The package
provides ```kform``` objects which are inherently alternating; they are described using  wedge products which are discussed next.

## Wedge products and the exterior calculus

This section follows the exposition of Hubbard and Hubbard, who
introduce the exterior calculus starting with a discussion of
elementary forms.  An elementary form is an object such as $dx_1\wedge
dx_3$, which is an alternating linear map from
$\mathbb{R}^n\times\mathbb{R}^n$ to $\mathbb{R}$ with

\[
\left(
dx_1\wedge dx_3
\right)\left(
\begin{pmatrix}a_1\\a_3\\ \vdots\\ a_n\end{pmatrix},
\begin{pmatrix}b_1\\b_3\\ \vdots\\ b_n\end{pmatrix}
\right)=\mathrm{det}
\begin{pmatrix} a_1 & b_1 \\ a_3 & b_3\end{pmatrix}
=a_1b_3-a_3b_1
\]

That this is alternating follows from the properties of the
determinant.  Because such objects are linear, it is possible to
consider more complicated forms such as $dx_1\wedge dx_2 + 3
dx_2\wedge dx_3$ with

\[
\left(
dx_1\wedge dx_2 + 3dx_2\wedge dx_3
\right)\left(
\begin{pmatrix}a_1\\a_2\\ \vdots\\ a_n\end{pmatrix},
\begin{pmatrix}b_1\\b_2\\ \vdots\\ b_n\end{pmatrix}
\right)=\mathrm{det}
\begin{pmatrix} a_1 & b_1\\ a_2 & b_2\end{pmatrix}
+3\mathrm{det}
\begin{pmatrix} a_2 & b_2\\ a_3 & b_3\end{pmatrix}
\]

or even $dx_1\wedge dx_2\wedge dx_3 +5dx_1\wedge dx_2\wedge dx_4$
which would be a linear map from $\left(\mathbb{R}^n\right)^3$ to
$\mathbb{R}$ with


\[
\left(
dx_4\wedge dx_2\wedge dx_3 +5dx_1\wedge dx_2\wedge dx_4
\right)\left(
\begin{pmatrix}a_1\\a_2\\ \vdots\\ a_n\end{pmatrix},
\begin{pmatrix}b_1\\b_2\\ \vdots\\ b_n\end{pmatrix},
\begin{pmatrix}c_1\\c_2\\ \vdots\\ c_n\end{pmatrix}
\right)=\mathrm{det}
\begin{pmatrix}
	a_4 & b_4 & c_4\\
	a_2 & b_2 & c_2\\
	a_3 & b_3 & c_3
\end{pmatrix}
+5\mathrm{det}
\begin{pmatrix}
	a_1 & b_1 & c_1\\
	a_2 & b_2 & c_2\\
	a_4 & b_4 & c_4
\end{pmatrix}.
\]

Taking the last form as an example, this has ready R idiom:

```{r}
M <- matrix(c(4,2,3,1,2,4),2,3,byrow=TRUE)
M
K <- as.kform(M,c(1,5))
K
```

Note that the order of the rows in ```K``` is immaterial and indeed in
some implementations will appear in a different order: the wedge
package uses the spray package, which in turn utilises the STL map
class of C++.

Spivak observes that $\Lambda^k(V)$ is spanned by the $n\choose k$
wedge products of the form

\[
dx_{i_1}\wedge dx_{i_2}\wedge\ldots\wedge dx_{i_k}\qquad
1\leq i_i<i_2<\cdots <i_k\leq n
\]

and these products are called elementary forms: every element of the
space $\Lambda^k(V)$ is a linear combination of elementary forms, as
illustrated in the package by function ```kform_general()```.
Consider the following idiom:


```{r}
Krel <- kform_general(4,2,1:6)
Krel
```

Object ```Krel``` is a two-form, specifically a map from
$\left(\mathbb{R}^4\right)^2$ to $\mathbb{R}$.  Observe that
```Krel``` has ${4\choose 2}=6$ components, which do not appear in any
particular order.

## Formal definition of wedge product

Given two alternating forms $\omega,\eta$, Spivak defines the wedge
product $\omega\wedge\eta\in\Lambda^{k+l}(V)$ as

\[
\omega\wedge\eta={k+l\choose k\,l}\operatorname{Alt}(\omega\otimes\eta)
\]

and this is implemented in the package by function ```wedge()```, or,
more idiomatically, ```%^%```:

```{r}
M1 <- matrix(c(3,4,5, 4,6,1),2,3,byrow=TRUE)
K1 <- as.kform(M1,c(2,7))
K1
M2 <- cbind(1:5,3:7)
K2 <- as.kform(M2,1:5)
K2
K1 %^% K2

```

See how the wedge product eliminates rows with repeated entries,
gathers permuted rows together (respecting the sign of the
permutation), and expresses the result in terms of elementary forms.
The product is a linear combination of two elementary forms but only
two coefficients are nonzero out of a possible ${7\choose 5}=21$
terms.  Note again that the order of the rows in the product is
arbitrary.

The wedge product has formal properties such as distributivity but by
far the most interesting one is associativity, which I will
demonstrate below:

```{r}
F1 <- as.kform(matrix(c(3,4,5, 4,6,1,3,2,1),3,3,byrow=TRUE))
F2 <- as.kform(cbind(1:6,3:8),1:6)
F3 <- kform_general(1:8,2)
(F1 %^% F2) %^% F3
F1 %^% (F2 %^% F3)
```


Note carefully in the above that the terms in ```(F1 %^% F2) %^% F3```
and ```F1 %^% (F2 %^% F3)``` appear in a different order.  They are
nevertheless algebraically identical:


```{r}
(F1 %^% F2) %^% F3 - F1 %^% (F2 %^% F3)
```


## Exterior derivatives

Hubbard and Hubbard define the exterior derivative $d\phi$ (they use a
bold font, $\mathbf{d}\phi$) of the $k$-form $\phi$ as the
$(k+1)$-form given by

\[  {d}\phi
  \left({v}_i,\ldots,{v}_{k+1}\right)
  =
  \lim_{h\longrightarrow 0}\frac{1}{h^{k+1}}\int_{\partial
  P_{x}\left(h{v}_1,\ldots,h{v}_{k+1}\right)}\phi
\]

which, by their own account, is a rather opaque mathematical idiom.
However, the definition makes sense and it can be shown that

\[
    {d}\left(f\,dx_{i_1}\wedge\cdots\wedge dx_{i_k}\right)=
    {d}f\wedge dx_{i_1}\wedge\cdots\wedge dx_{i_k}
\]

where $f\colon\mathbb{R}^n\longrightarrow\mathbb{R}$ is a scalar
function of position.  The package provides ```grad()```, which when
given a vector $x_1,\ldots,x_n$ returns the one-form


\[
\sum_{i=1}^n x_idx_i
\]

This is useful because $df=\sum_{j=1}^n\left(D_j f\right)\,dx_j$. Thus

```{r}
grad(c(0.4,0.1,-3.2,1.5))
```

We will use the ```grad()``` function to verify that, in
$\mathbb{R}^n$, a certain $(k-1)$-form has zero work function.
Motivated by the fact that

\[
F_3=\frac{1}{\left(x^2+y^2+z^2\right)^{3/2}}
\begin{pmatrix}x\\y\\z\end{pmatrix}
\]

is a divergenceless velocity field in $\mathbb{R}^3$, H\&H go on to define

\[
\omega_n=
d\frac{1}{\left(x_1^2+\ldots +x_n^2\right){^n/2}}\sum_{i=1}^{n}(-1)^{i-1}
x_idx_1\wedge\cdots\wedge\widehat{dx_i}\wedge\cdots\wedge dx_n
\]

(where a hat indicates the absence of a term), and show analyically
that $d\omega=0$.  Here I show this using R idiom.  The first thing is
to define a function that implements the hat:


```{r}
f <- function(x){
    n <- length(x)
    as.kform(t(apply(diag(n)<1,2,which)))
}
```

So, for example:

```{r}
f(1:5)
```

Then we can use the ```grad()``` function to calculate $d\omega$,
using the quotient law to express the derivatives analytically:

```{r}
df  <- function(x){
    n <- length(x)
    S <- sum(x^2)
    grad(rep(c(1,-1),length=n)*(S^(n/2) - n*x^2*S^(n/2-1))/S^n
    )
}
```

Thus
```{r}
df(1:5)
```

Now we can use the wedge product of the two parts to show that the exterior derivative is zero:

```{r}
x <- rnorm(9)
print(df(x) %^% f(x))  # should be zero
```

# Differential of the differential, $d^2=0$

We can use the package to verify the celebrated fact that, for any
$k$-form $\phi$, $d\left(d\phi\right)=0$.  The first step is to define
scalar functions ```f1(), f2(), f3()```, all  $0$-forms:

```{r}
f1 <- function(w,x,y,z){w*x*y*z + sin(x) + cos(w)}
f2 <- function(w,x,y,z){w^2*x*y*z + sin(w) + w+z}
f3 <- function(w,x,y,z){w*y + x + (w*z)}
```

Now we need to define elementary $1$-forms:

```{r}
dw <- as.kform(1)
dx <- as.kform(2)
dy <- as.kform(3)
dz <- as.kform(4)
```


I will demonstrate the theorem by defining a $2$-form which is the sum
of three elementary two-forms, evaluated at a particular point in
$\mathbb{R}^4$:

```{r}
phi <-
  (
    +f1(1,2,3,4) %^% dw %^% dx
    +f2(1,2,3,4) %^% dw %^% dy
    +f3(1,2,3,4) %^% dy %^% dz
  )
```

We can use slightly slicker R idiom by defining elementary forms
```e1,e2,e3``` and then defining phi to be a linear sum, weighted with
$0$-forms given by the (scalar) functions:

```{r}
e1 <- dw %^% dx
e2 <- dw %^% dy
e3 <- dy %^% dz

phi <-
  (
    +f1(1,2,3,4) %^% e1
    +f2(1,2,3,4) %^% e2
    +f3(1,2,3,4) %^% e3
  )
phi
```


Now to evaluate first derivatives of ```f1()``` etc at point $(1,2,3,4)$, using ```Deriv()```:

```{r}
Df1 <- Deriv(f1)(1,2,3,4)
Df2 <- Deriv(f2)(1,2,3,4)
Df3 <- Deriv(f3)(1,2,3,4)
```

So ```Df1``` etc are numeric vectors of length 4.  Calculating dphi is easy:

```
dphi <-
  (
    +grad(Df1) %^% e1
    +grad(Df2) %^% e2
    +grad(Df3) %^% e3
  )
```

Now work on the differential of the differential.  First evaluate
the Hessians (4x4 numeric matrices) at the same point:

```{r}
Hf1 <- matrix(Deriv(f1,nderiv=2)(1,2,3,4),4,4)
Hf2 <- matrix(Deriv(f2,nderiv=2)(1,2,3,4),4,4)
Hf3 <- matrix(Deriv(f3,nderiv=2)(1,2,3,4),4,4)
```

But $dd\phi$ is clearly zero as the Hessians are symmetrical:
```{r}
ddphi <- # should be zero
  (  
    +as.kform(which(!is.na(Hf1),arr.ind=TRUE),c(Hf1))
    +as.kform(which(!is.na(Hf2),arr.ind=TRUE),c(Hf2))
    +as.kform(which(!is.na(Hf3),arr.ind=TRUE),c(Hf3))
  )

ddphi
```

as expected.

## Stokes's theorem

In its most general form, Stokes's theorem states

\[
\int_{\partial X}\phi=\int_Xd\phi
\]

where $X\subset\mathbb{R}^n$ is a compact oriented $(k+1)$-dimensional
manifold with boundary $\partial X$ and $\phi$ is a $k$-form defined
on a neighborhood of $X$.

We will verify Stokes, following 6.9.5 of Hubbard in which

\[
\phi=
\left(x_1-x_2^2+x_3^3-\cdots\pm x_n^n\right)
\left(
\sum_{i=1}^n
dx_1\wedge\cdots\wedge\widehat{dx_i}\wedge\cdots\wedge dx_n
\right)
\]

(a hat indicates that a term is absent), and we wish to evaluate
$\int_{\partial C_a}\phi$ where $C_a$ is the cube $0\leq x_j\leq a,
1\leq j\leq n$.  Stokes tells us that this is equal to $\int_{C_a}
d\phi$, which is given by

\[
d\phi = \left(
1+2x_2+\cdots + nx_n^{n-1}\right)
dx_1\wedge\cdots\wedge dx_n
\]

and so the volume integral is just

\[
\sum_{j=1}^n
\int_{x_1=0}^a
\int_{x_2=0}^a
\cdots
\int_{x_i=0}^a
jx_j^{j-1}
dx_1 dx_2\ldots dx_n=
a^{n-1}\left(a+a^2+\cdots+a^n\right).
\]

Stokes's theorem, being trivial, is not amenable to direct numerical
verification but the package does allow slick creation of $\phi$:

```{r}
phi <- function(x){
    n <- length(x)
    sum(x^seq_len(n)*rep_len(c(1,-1),n)) * as.kform(t(apply(diag(n)<1,2,which)))
}
phi(1:9)
```

Further, $d\phi$ is given by

```{r}
dphi <- function(x){
    nn <- seq_along(x)
    sum(nn*x^(nn-1)) * as.kform(seq_along(x))
}
dphi(1:9)
```

## References

*  J. H. Hubbard and B. B. Hubbard 2015.  _Vector
      calculus, linear algebra and differential forms: a unified
      aproach_,  Ithaca, NY.
*  M. Spivak 1971. _Calculus on manifolds_, Addison-Wesley.
