---
title: "Exterior calculus with R"
author: "Robin K. S. Hankin"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{The exterior calculus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# The exterior calculus

Ordinary differential calculus may be generalized to
arbitrary-dimensional oriented manifolds using the exterior calculus.
Here I show how the wedge package furnishes functionality for working
with the exterior calculus, and provide numerical verification of a
number of theorems.

Notation follows that of Spivak, and Hubbard and Hubbard.

```{r,echo=FALSE,print=FALSE,results='hide'}
require("wedge",quietly=TRUE,warn.conflicts=FALSE)
```

Recall that a $k$-tensor is a multilinear map $S\colon
V^k\longrightarrow\mathbb{R}$, where $V=\mathbb{R}^n$ is considered as
a vector space; Spivak denotes the space of multilinear maps as
$\mathcal{J}^k(V)$.  Formally, multilinearity means

\[
S\left(v_1,\ldots,av_i,\ldots,v_k\right)=a\cdot S\left(v_1,\ldots,v_i,\ldots,v_k\right)
\]

and

\[S\left(v_1,\ldots,v_i+{v_i}',\ldots,v_k\right)=S\left(v_1,\ldots,v_i,\ldots,x_v\right)+
S\left(v_1,\ldots,{v_i}',\ldots,v_k\right).
\]

where $v_i\in V$.  If $S\in\mathcal{J}^k(V)$ and
$T\in\mathcal{J}^l(V)$, then we may define $S\otimes
T\in\mathcal{J}^{k+l}(V)$ as

\[
S\otimes T\left(v_1,\ldots,v_k,v_{k+1},\ldots,v_{k+l}\right)=
S\left(v_1,\ldots,v_k\right)\cdot
T\left(v_1,\ldots,v_l\right)
\]


Spivak observes that $\mathcal{J}^k(V)$ is spanned by the $n^k$
products of the form

\[
\phi_{i_1}\otimes\phi_{i_2}\otimes\ldots\otimes\phi_{i_k}\qquad
1\leq i_i,i_2,\ldots,i_k\leq n
\]

where $v_1,\ldots,v_k$ is a basis for $V$ and
$\phi_i\left(v_j\right)=\delta_{ij}$; we can therefore write

\[
S=\sum_{1\leq i_1,\ldots,i_k\leq n} a_{i_1\ldots i_k} 
\phi_{i_1}\otimes\ldots\otimes\phi_{i_k}.
\]

The space spanned by such products has a natural representation in R
as an array of dimensions $n\times\cdots\times n$.  If ```A``` is such
an array, then the element ```A[i_1,i_2,...,i_k]``` is the coefficient
of $\phi_{i_1}\otimes\ldots\otimes\phi_{i_k}$.  However, it is more
efficient and conceptually cleaner to consider a *sparse* array.
Considering the case $n=5,k=4$

We can test these identities in R using the idiom furnished by the
wedge package:

```{r}
k <- 4
n <- 5
M <- matrix(c(5,1,1,1, 1,1,2,3, 1,3,4,2),3,4,byrow=TRUE)
M
S <- as.ktensor(M,coeffs= 0.5 + 1:3)
S
```

So $S$ is a 4-tensor, mapping $V^4$ to $\mathbb{R}$, where
$V=\mathbb{R}^5$.  Here we have
$S=1.5\phi_5\otimes\phi_1\otimes\phi_1\otimes\phi_1+2.5\phi_1\otimes\phi_1\otimes\phi_2\otimes\phi_3+3.5\phi_1\otimes\phi_3\otimes\phi_4\otimes\phi_2$.
Note that on some implementations the row order of object ```S``` will
differ from that of ```M```.  We will define $E$ to be a random point
in $V^k$ in terms of a matrix:

```{r}
E <- matrix(rnorm(n*k),n,k)   # A random point in V^k
```

Then

```{r}
f <- as.function(S)
f(E)
```

Testing  multilinearity is straightforward in the package:
```{r}

E1 <- E
E2 <- E
E3 <- E

x1 <- rnorm(n)
x2 <- rnorm(n)
r1 <- rnorm(1)
r2 <- rnorm(1)

E1[,2] <- x1
E2[,2] <- x2
E3[,2] <- r1*x1 + r2*x2
```

Then we can verify the multilinearity of $S$:

```{r}

f <- as.function(S)
c(r1*f(E1) + r2*f(E2), f(E3)) # should match
```


Note that this is *not* equivalent to linearity over $V^{nk}$:


```{r}
E1 <- matrix(rnorm(n*k),n,k)
E2 <- matrix(rnorm(n*k),n,k)
c(f(E1+E2),f(E1)+f(E2))
```

Given two k-tensor objects \eqn{S,T} we can form the cross product
$S\otimes T$, defined as

\[
S\otimes T\left(v_1,\ldots,v_k,v_{k+1},\ldots, v_{k+l}\right)=
  S\left(v_1,\ldots v_k\right)\cdot T\left(v_{k+1},\ldots
  v_{k+l}\right)
\]


```{r}
S1 <- ktensor(spray(cbind(1:3,2:4),1:3))
S2 <- as.ktensor(matrix(1:6,2,3))
S1
S2
```

The R idiom for $S1\otimes S2$ would be ```cross()```, or ```%X%```:


```{r}
cross(S1,S2)
```

Then, for example:

```{r}
as.function(cross(S1,S2))(matrix(rnorm(30),6,5))
```

# Alternating forms

An alternating form is a multilinear map $T$ satisfying

\[
\mathrm{T}\left(v_1,\ldots,v_i,\ldots,v_j,\ldots,v_k\right)=
    -\mathrm{T}\left(v_1,\ldots,v_j,\ldots,v_i,\ldots,v_k\right)
\]

(or, equivalently,
$\mathrm{T}\left(v_1,\ldots,v_i,\ldots,v_i,\ldots,v_k\right)= 0$).  We
write $\Lambda^k(V)$ for the space of all alternating multilinear maps
from $V^k$ to $\mathbb{R}$.  Spivak gives
$\operatorname{Alt}\colon\mathcal{J}^k(V)\longrightarrow\Lambda^k(V)$
defined by

\[\operatorname{Alt}(T)\left(v_1,\ldots,v_k\right)=
    \frac{1}{k!}\sum_{\sigma\in S_k}\operatorname{sgn}(\sigma)\cdot
    T\left(v_{\sigma(1)},\ldots,v_{\sigma(k)}\right)
\]

where the sum ranges over all permutations of
$\left[n\right]=\left\{1,2,\ldots,n\right\}$ and
$\operatorname{sgn}(\sigma)\in\pm 1$ is the sign of the permutation.
If $T\in\mathcal{J}^k(V)$ and $\omega\in\Lambda^k(V)$, it is
straightforward to prove that $\operatorname{Alt}(T)\in\Lambda^k(V)$,
$\operatorname{Alt}\left(\operatorname{Alt}\left(T\right)\right)=\operatorname{Alt}\left(T\right)$,
and $\operatorname{Alt}\left(\omega\right)=\omega$.

In the wedge package, this is effected by the ```Alt()``` function:


```{r}
S1
Alt(S1)
```

And verifying that it is in fact alternating is straightforward:

```{r}
E <- matrix(rnorm(8),4,2)
Erev <- E[,2:1]
as.function(Alt(S1))(E) + as.function(Alt(S1))(Erev)  # should be zero
```

However, we can see that this form for alternating tensors (here
called $k$-forms) is inefficient and highly redundant.  The package
provides ```kform``` objects which are inherently alternating; they are described using  wedge products which are discussed next.

## Wedge products and the exterior calculus

This section follows the exposition of Hubbard and Hubbard, who
introduce the exterior calculus starting with a discussion of
elementary forms.  An elementary form is an object such as $dx_1\wedge
dx_3$, which is an alternating linear map from
$\mathbb{R}^n\times\mathbb{R}^n$ to $\mathbb{R}$ with

\[
\left(
dx_1\wedge dx_3
\right)\left(
\begin{pmatrix}a_1\\a_3\\ \vdots\\ a_n\end{pmatrix},
\begin{pmatrix}b_1\\b_3\\ \vdots\\ b_n\end{pmatrix}
\right)=\mathrm{det}
\begin{pmatrix} a_1 & b_1 \\ a_3 & b_3\end{pmatrix}
=a_1b_3-a_3b_1
\]

That this is alternating follows from the properties of the
determinant.  Because such objects are linear, it is possible to
consider more complicated forms such as $dx_1\wedge dx_2 + 3
dx_2\wedge dx_3$ with

\[
\left(
dx_1\wedge dx_2 + 3dx_2\wedge dx_3
\right)\left(
\begin{pmatrix}a_1\\a_2\\ \vdots\\ a_n\end{pmatrix},
\begin{pmatrix}b_1\\b_2\\ \vdots\\ b_n\end{pmatrix}
\right)=\mathrm{det}
\begin{pmatrix} a_1 & b_1\\ a_2 & b_2\end{pmatrix}
+3\mathrm{det}
\begin{pmatrix} a_2 & b_2\\ a_3 & b_3\end{pmatrix}
\]

or even $dx_1\wedge dx_2\wedge dx_3 +5dx_1\wedge dx_2\wedge dx_4$
which would be a linear map from $\left(\mathbb{R}^n\right)^3$ to
$\mathbb{R}$ with


\[
\left(
dx_4\wedge dx_2\wedge dx_3 +5dx_1\wedge dx_2\wedge dx_4
\right)\left(
\begin{pmatrix}a_1\\a_2\\ \vdots\\ a_n\end{pmatrix},
\begin{pmatrix}b_1\\b_2\\ \vdots\\ b_n\end{pmatrix},
\begin{pmatrix}c_1\\c_2\\ \vdots\\ c_n\end{pmatrix}
\right)=\mathrm{det}
\begin{pmatrix}
	a_4 & b_4 & c_4\\
	a_2 & b_2 & c_2\\
	a_3 & b_3 & c_3
\end{pmatrix}
+5\mathrm{det}
\begin{pmatrix}
	a_1 & b_1 & c_1\\
	a_2 & b_2 & c_2\\
	a_4 & b_4 & c_4
\end{pmatrix}.
\]

Taking the last form as an example, this has ready R idiom:

```{r}
M <- matrix(c(4,2,3,1,2,4),2,3,byrow=TRUE)
M
K <- as.kform(M,c(1,5))
K
```

Note that the order of the rows in ```K``` is immaterial and indeed in
some implementations will appear in a different order: the wedge
package uses the spray package, which in turn utilises the STL map
class of C++.

Spivak observes that $\Lambda^k(V)$ is spanned by the $n\choose k$
wedge products of the form

\[
dx_{i_1}\wedge dx_{i_2}\wedge\ldots\wedge dx_{i_k}\qquad
1\leq i_i<i_2<\cdots <i_k\leq n
\]

and these products are called elementary forms: every element of the
space $\Lambda^k(V)$ is a linear combination of elementary forms, as
illustrated in the package by function ```kform_general()```.
Consider the following idiom:


```{r}
Krel <- kform_general(4,2,1:6)
Krel
```

Object ```Krel``` is a two-form, specifically a map from
$\left(\mathbb{R}^4\right)^2$ to $\mathbb{R}$.  Observe that
```Krel``` has ${4\choose 2}=6$ components, which do not appear in any
particular order.

## Formal definition of wedge product

Given two alternating forms $\omega,\eta$, Spivak defines the wedge
product $\omega\wedge\eta\in\Lambda^{k+l}(V)$ as

\[
\omega\wedge\eta={k+l\choose k\,l}\operatorname{Alt}(\omega\otimes\eta)
\]

and this is implemented in the package by function ```wedge()```, or,
more idiomatically, ```%^%```:

```{r}
M1 <- matrix(c(3,4,5, 4,6,1),2,3,byrow=TRUE)
K1 <- as.kform(M1,c(2,7))
K1
M2 <- cbind(1:5,3:7)
K2 <- as.kform(M2,1:5)
K2
K1 %^% K2

```

See how the wedge product eliminates rows with repeated entries,
gathers permuted rows together (respecting the sign of the
permutation), and expresses the result in terms of elementary forms.
The product is a linear combination of two elementary forms but only
two coefficients are nonzero out of a possible ${7\choose 5}=21$
terms.  Note again that the order of the rows in the product is
arbitrary.

The wedge product has formal properties such as distributivity but by
far the most interesting one is associativity, which I will
demonstrate below:

```{r}
F1 <- as.kform(matrix(c(3,4,5, 4,6,1,3,2,1),3,3,byrow=TRUE))
F2 <- as.kform(cbind(1:6,3:8),1:6)
F3 <- kform_general(1:8,2)
(F1 %^% F2) %^% F3
F1 %^% (F2 %^% F3)
```


Note carefully in the above that the terms in ```(F1 %^% F2) %^% F3```
and ```F1 %^% (F2 %^% F3)``` appear in a different order.  They are
nevertheless algebraically identical:


```{r}
(F1 %^% F2) %^% F3 - F1 %^% (F2 %^% F3)
```
